\documentclass[11pt, a4paper]{article}
\input{preamble}

\renewcommand{\doccoursename}{Neural Networks for Engineers}
\renewcommand{\docsessionname}{Session 5 — Quiz}

\begin{document}

\maketitleblock{Session 5 — Quiz}{Loss Functions \& Gradient Descent \quad|\quad 10 questions}{mainblue}

\renewcommand{\sectionprefix}{Question}

% ------------------------------------------------------------------
\section{MSE by Hand}

A model produces the following predictions. Compute the MSE.

\begin{center}
\begin{tabular}{cccc}
\toprule
Sample & True $y$ & Predicted $\hat{y}$ & Squared Error \\
\midrule
1 & 2.0 & 2.5 & \acell \\
2 & 4.0 & 3.0 & \acell \\
3 & 1.0 & 1.5 & \acell \\
\bottomrule
\end{tabular}
\end{center}

\vspace{0.5em}
\textbf{MSE} $=$ \acell

% ------------------------------------------------------------------
\section{Why Squaring?}

Why do we square the errors in MSE rather than simply averaging them?

\vspace{5em}

% ------------------------------------------------------------------
\section{Derivative of a Loss}

Given $L(w) = (4 - 2w)^2$, compute $\dfrac{dL}{dw}$.

\vspace{5em}

% ------------------------------------------------------------------
\section{Gradient Direction}

You are at $w = 0$ on the loss curve of Question~3. The gradient is negative.
Should you \textbf{increase} or \textbf{decrease} $w$ to reduce the loss? Explain why.

\vspace{5em}

% ------------------------------------------------------------------
\section{Manual Gradient Descent}

Perform \textbf{2 steps} of gradient descent to minimize $L(w) = w^2 - 6w + 10$,
starting at $w_0 = 0$ with learning rate $\eta = 0.2$.

\begin{center}
\begin{tabular}{ccccc}
\toprule
Step & $w$ & $L(w)$ & $\dfrac{dL}{dw}$ & New $w$ \\
\midrule
0 & 0 & \acell & \acell & \acell \\[1.2em]
1 & \acell & \acell & \acell & \acell \\[1.2em]
\bottomrule
\end{tabular}
\end{center}

% ------------------------------------------------------------------
\section{Loss Landscape}

What is the difference between a \textbf{local minimum} and a \textbf{global minimum}
on a loss landscape? For which type of model is the MSE loss guaranteed to have
only one minimum?

\vspace{6em}

% ------------------------------------------------------------------
\section{Partial Derivatives}

Given $L(w_1, w_2) = (2 - w_1 - 3w_2)^2$, compute the gradient $\nabla L$
and evaluate it at $(w_1, w_2) = (1,\, 0)$.

\vspace{6em}

% ------------------------------------------------------------------
\section{SGD Variants}

Complete the table.

\begin{center}
\begin{tabularx}{\linewidth}{lccX}
\toprule
Variant & Batch size & Gradient estimate & Typical use case \\
\midrule
Full-batch GD   & $N$        & \acell & \acell \\[1em]
Mini-batch SGD  & $B \ll N$  & \acell & \acell \\[1em]
Stochastic GD   & \acell     & \acell & \acell \\[1em]
\bottomrule
\end{tabularx}
\end{center}

% ------------------------------------------------------------------
\section{Diverging Loss}

A colleague trains a linear regression model and observes that the loss
\textbf{increases} over the first few epochs. What is the most likely cause,
and how would you fix it?

\vspace{6em}

% ------------------------------------------------------------------
\section{Bridge to Neural Networks}

The notebook argues that gradient descent on linear regression and on a
multi-layer neural network follow the \textbf{same recipe}, with one key
difference. What is that difference, and what technique (covered in Session~6)
addresses it?

\vspace{6em}

\end{document}
